---
title: "Analysing GC Content with kirgcdemo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing GC Content with kirgcdemo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Introduction

The `kirgcdemo` package provides two core utilities for DNA sequence analysis:

- `gc_content()` — percentage of G/C bases in a sequence
- `reverse_complement()` — Watson–Crick reverse complement

These are deliberately simple functions, chosen to keep the packaging concepts
centre-stage. The same packaging techniques apply to much more complex tools.

```{r library}
library(kirgcdemo)
```

---

## GC Content Across a Set of Sequences

GC content varies widely across genomes and is biologically meaningful —
high-GC regions are often gene-dense and thermally more stable. Promoter
regions tend to be AT-rich; exons are often GC-enriched.

```{r gc-example}
seqs <- c(
  promoter = "ATATATATATAT",
  exon     = "GCGCGCGCATGC",
  mixed    = "ATGCATGCATGC"
)

# Native pipe |> (R ≥ 4.1): no magrittr required
gc_values <- seqs |> vapply(gc_content, numeric(1))
gc_values
```

```{r barplot, fig.width=5, fig.height=4}
barplot(
  gc_values,
  ylab = "GC content (%)",
  main = "GC content by region",
  col  = c("#4E9AF1", "#F1714E", "#6DBF67"),
  ylim = c(0, 100)
)
abline(h = 50, lty = 2, col = "grey50")
```

---

## Reverse Complement

The reverse complement is fundamental to working with double-stranded DNA —
primers, probes, and RNA-seq adapters are all defined relative to one strand.

```{r revcomp-example}
seq <- "ATGCTAGCTAGCT"
cat("Forward:            ", seq, "\n")
cat("Reverse complement: ", reverse_complement(seq), "\n")
```

A useful property: the reverse complement of the reverse complement is the
original sequence.

```{r identity-check}
original <- "GCTAGCTAGCTA"
stopifnot(reverse_complement(reverse_complement(original)) == original)
cat("Round-trip check passed.\n")
```

---

## Handling Ambiguous Bases

Real sequencing data often contains `N` (any base). `gc_content()` counts `N`
in the denominator but not as GC, diluting the result:

```{r ambiguous}
gc_content("GCNN")   # 50% — two GC out of four total bases
gc_content("GGNN")   # 50% — two G out of four total bases
gc_content("GGGG")   # 100% — no ambiguity
```

`reverse_complement()` passes ambiguous bases through unchanged:

```{r ambiguous-revcomp}
reverse_complement("ATGCN")   # "NGCAT"
```
